An open and continuously updated fern tree of life (FTOL)

### 2022-02-13 ms/manuscript.Rmd
Calculate family stem ages

tar_load(
	c(
		sanger_alignment,
		plastome_alignment,
		plastome_metadata_renamed,
		ppgi_taxonomy,
		plastid_calibration_dates,
		pteridocat,
		sanger_alignment_tbl,
		sanger_tree_fast,
		sanger_sampling,
		monophy_by_clade,
		plastome_tree,
		plastid_tree_dated,
		fossil_calibration_tips,
		treepl_cv_results),
	store = ftol_cache
)

@@

# Make tibble with one row per species in Sanger sampling
sanger_sampling <- tibble(
	species = rownames(sanger_alignment)
) %>%
	# Add higher-level taxonomy
	mutate(genus = str_split(species, "_") %>% map_chr(1)) %>%
	left_join(
		select(ppgi_taxonomy, genus, family, suborder, order), by = "genus"
	) %>% 
	# Add plot_group
	add_plot_group() %>%
	# Add outgroup status
	left_join(og_species, by = "species") %>%
	mutate(outgroup = replace_na(outgroup, FALSE)) %>%
	verify(nrow(.) == nrow(sanger_alignment)) %>%
	verify(sum(outgroup) == nrow(og_species))

@@

```{r divergence-times-stats}
# Count number of tips in clades by taxonomy for double checking
n_tips_in_clades <-
	sanger_sampling %>%
	filter(outgroup == FALSE) %>%
	select(species, family, order) %>%
	pivot_longer(names_to = "level", values_to = "taxon", -species) %>%
	group_by(taxon) %>%
	summarize(number_tips_by_tax = n())

tree_height <- max(phytools::nodeHeights(plastid_tree_dated))

@@

# Get ages for large clades
clade_ages <-
	sanger_sampling %>%
	select(family, order) %>%
	unique() %>%
	pivot_longer(names_to = "level", values_to = "taxon", everything()) %>%
	filter(!is.na(taxon)) %>%
	unique() %>%
	left_join(
		unique(select(monophy_by_clade, taxon, monophyly, mrca, number_tips)),
		by = "taxon") %>%
	mutate(across(c(mrca, number_tips), parse_number)) %>%
	assert(in_set(c("Yes", "Monotypic")), monophyly) %>%
	filter(monophyly == "Yes") %>%
	mutate(
		# Height is distance above root
		height = map_dbl(mrca, ~phytools::nodeheight(
			tree = plastid_tree_dated, node = .))
	) %>%
	mutate(
		# Age is the total length of the tree minus height
		age = tree_height - height
	) %>%
	# Double check that n_tips is as expected
	# (so the correct node is being used)
	left_join(n_tips_in_clades, by = "taxon") %>%
	verify(all(number_tips == number_tips_by_tax)) %>%
	select(-number_tips_by_tax)
```

@@

```{r div-times-make}
# Convert family ages to long format
family_ages_median <-
clade_ages %>%
	filter(level == "family") %>%
	select(family = taxon, current_best = age) %>%
	left_join(other_dates, by = "family") %>%
	mutate(family = fct_reorder(family, current_best)) %>%
	pivot_longer(values_to = "age", -family) %>%
	separate(name, into = c("author", "age_type"), sep = "_") %>%
	assert(not_na, author, age_type) %>%
	filter(age_type %in% c("best", "median"))

# Convert family errors to almost long format
family_ages_errors <-
	clade_ages %>%
	filter(level == "family") %>%
	select(family = taxon, current_best = age) %>%
	left_join(other_dates, by = "family") %>%
	mutate(family = fct_reorder(family, current_best)) %>%
	select(family, matches("ts|rothfels")) %>%
	select(-matches("median"))

>>>>

tar_load(
	c(
		sanger_alignment,
		plastome_alignment,
		plastome_metadata_renamed,
		ppgi_taxonomy,
		plastid_calibration_dates,
		pteridocat,
		sanger_alignment_tbl,
		sanger_tree_fast,
		monophy_by_clade,
		plastome_tree,
		plastid_tree_dated,
		fossil_calibration_tips,
		treepl_cv_results),
	store = ftol_cache
)

@@

# Make tibble with one row per species in Sanger sampling
sanger_sampling <- make_sanger_sampling_tbl(
	plastome_metadata_renamed,
	sanger_tree = plastid_tree_dated,
	ppgi_taxonomy = ppgi_taxonomy)

@@

```{r divergence-times-stats}
# Get overall tree height
tree_height <- max(phytools::nodeHeights(plastid_tree_dated))

@@

# Get ages for families
sanger_sampling_family <- select(sanger_sampling, species, family)

# Only use monophyletic families
mono_test_family <- assess_monophy(
	taxon_sampling = sanger_sampling_family,
	tree = plastid_tree_dated,
	tax_levels = "family"
) %>%
	get_result_monophy(1)

family_ages <-
mono_test_family %>%
	filter(monophyly == "Yes") %>%
	mutate(
		mrca = parse_number(mrca),
		stem_node = map_dbl(
			mrca,
			~phangorn::Ancestors(plastid_tree_dated, ., type = "parent"))
	) %>%
	select(family = taxon, stem_node) %>%
mutate(
	# Height is distance above root
	height = map_dbl(stem_node, ~phytools::nodeheight(
		tree = plastid_tree_dated, node = .))
) %>%
	mutate(
		# Age is the total length of the tree minus height
		age = tree_height - height
	)
```

@@

```{r div-times-make}
# Convert family ages to long format
family_ages_median <-
family_ages %>%
	select(family, current_best = age) %>%
	left_join(other_dates, by = "family") %>%
	mutate(family = fct_reorder(family, current_best)) %>%
	pivot_longer(values_to = "age", -family) %>%
	separate(name, into = c("author", "age_type"), sep = "_") %>%
	assert(not_na, author, age_type) %>%
	filter(age_type %in% c("best", "median"))

# Convert family errors to almost long format
family_ages_errors <-
	family_ages %>%
	select(family, current_best = age) %>%
	left_join(other_dates, by = "family") %>%
	mutate(family = fct_reorder(family, current_best)) %>%
	select(family, matches("ts|rothfels")) %>%
	select(-matches("median"))
